Clustering: A pipeline to analyse time-course gene expression data

### 2019-03-05 scripts/01_quality_control_normalization.Rmd
MAINT removing a lot of unused packages

```{r echo=FALSE, results="hide"}
# Loading dependencies
library(devtools)
library(limma)
library(splines)
library(stats)
library(moanin)
source("utils.R")

@@

library(genefilter)
library(dplyr)
library(ggplot2)
library(NMF)
library(reshape2)
library(limma)
```

@@

# Preform background correction using norm-exponential strategy and normalize
# across samples using quantile normalization.
RG = backgroundCorrect(RG, method="normexp")
RG.q = normalizeBetweenArrays(RG, method="quantile")
```

@@

```{r plotting_densities, results="hide", fig.width=8, fig.height=4}
# Plot the densities of the probes before and after normalization.
par(mfrow=c(1, 2))
plotDensities(RG, col="black", legend=FALSE)
plotDensities(RG.q, col="black", legend=FALSE)
```

@@

```{r}
variance_cutoff = 0.5
variance_per_genes = apply(data, 1, mad)
min_variance = quantile(variance_per_genes, c(variance_cutoff))
data = data[variance_per_genes > min_variance,]
```

```{r echo=FALSE, fig.width=8, fig.height=4}
hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
     ylab="# probes")
abline(v=min_variance, col="#AB0000", lw=2)
```

@@

library(ggfortify)
pca_data = prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))
```

@@

```{r pca_plots, fig.width=8, fig.height=8}
# FIXME Add legend
par(mfrow=c(2, 2))
plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
     col=ann_colors$Group[meta$Group],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC1")
plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
     col=ann_colors$Time[as.factor(meta$Time)],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC1")

plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
     col=ann_colors$Group[meta$Group],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC3")
plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
     col=ann_colors$Time[as.factor(meta$Time)],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
     xlab="PC2", ylab="PC3")
```

@@

data_corr = cor(data, method="pearson")

meta$Time = as.factor(meta$Time)
aheatmap(data_corr,
         Colv=NA, Rowv=NA,
         annCol=meta[,
		     c("Group", "Time")],
         annLegend=FALSE, 
	 annColors=ann_colors,
         main="Correlation plot")
```

>>>>

```{r echo=FALSE, results="hide"}
# Loading dependencies
library(devtools)
library(limma)
library(splines)
library(stats)
library(moanin)
library(ggplot2)
library(stats)

source("utils.R")

@@

```

@@

# Preform background correction using norm-exponential strategy and normalize
# across samples using quantile normalization.
RG = limma::backgroundCorrect(RG, method="normexp")
RG.q = limma::normalizeBetweenArrays(RG, method="quantile")
```

@@

```{r plotting_densities, results="hide", fig.width=8, fig.height=4}
# Plot the densities of the probes before and after normalization.
par(mfrow=c(1, 2))
limma::plotDensities(RG, col="black", legend=FALSE)
limma::plotDensities(RG.q, col="black", legend=FALSE)
```

@@

```{r}
variance_cutoff = 0.5
variance_per_genes = apply(data, 1, mad)
min_variance = stats::quantile(variance_per_genes, c(variance_cutoff))
data = data[variance_per_genes > min_variance,]
```

```{r echo=FALSE, fig.width=8, fig.height=4}
graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
	       ylab="# probes")
graphics::abline(v=min_variance, col="#AB0000", lw=2)
```

@@

pca_data = stats::prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))
```

@@

```{r pca_plots, fig.width=8, fig.height=8}
# FIXME Add legend
graphics::par(mfrow=c(2, 2))
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Time[as.factor(meta$Time)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")

graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC3")
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
    col=ann_colors$Time[as.factor(meta$Time)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC3")
```

@@

data_corr = stats::cor(data, method="pearson")

meta$Time = as.factor(meta$Time)
NMF::aheatmap(
    data_corr,
    Colv=NA, Rowv=NA,
    annCol=meta[,
		c("Group", "Time")],
    annLegend=FALSE, 
    annColors=ann_colors,
    main="Correlation plot")
```

### 2019-03-05 scripts/03_clustering.Rmd
MAINT removing a lot of unused packages

# Right now, the package doesn't contain the normalized data
de_analysis = read.table(".results/pvalues.txt", sep="\t", check.names=FALSE)
```

@@

```{r filter_genes}
if(filter_genes){
    # First, filter out any genes that doesn't have a log fold change of at
    # least 2 at at least one of the time points.
    lfc_col_to_keep = colnames(de_analysis)[
	grepl("-lfc_max", colnames(de_analysis))]
    de_analysis = de_analysis[rowSums(abs(de_analysis[, lfc_col_to_keep]) > 2) > 0, ]

    if(dim(de_analysis)[1] > n_genes_to_keep){

	# Then rank by fisher's p-value and take max the number of genes of
	# interest
	# Filter out q-values for the pvalues table
	pval_col_to_keep = colnames(de_analysis)[grepl("-pval", colnames(de_analysis))]
	pvalues = de_analysis[, pval_col_to_keep]
	fishers_pval = moanin:::fisher_method(pvalues)
	fishers_qval = stats::p.adjust(fishers_pval)
        if(keep_all_significant){
	   genes_to_keep = names(fishers_qval[fishers_qval < 0.05])
	}else{
	    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
	}
    }else{
	genes_to_keep = names(pvalues)
    }

@@

# Rescale the centroids
centroids = moanin:::rescale_values(kmeans_clusters$centroids, meta)
all_scores = data.frame(row.names=row.names(data))
for(k in 1:n_clusters){
    scores = moanin:::score_genes_centroid(data, centroids[k,])
    all_scores[paste0("cluster_", k)] = scores
}

# Only assign labels to X% of the genes
max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
labels = moanin:::row_argmin(all_scores)
labels[genes_to_not_consider] = NA
```

@@

```{r}
table(labels)
```

@@

>>>>

# Right now, the package doesn't contain the normalized data
de_analysis = read.table(".results/pvalues.txt", sep="\t", check.names=FALSE)

# Extract pvalues and log_fold change
lfc_col_to_keep = colnames(de_analysis)[
    grepl("-lfc_max", colnames(de_analysis))]
lfc_max = de_analysis[, lfc_col_to_keep]

pval_col_to_keep = colnames(de_analysis)[
    grepl("-pval", colnames(de_analysis))]
pvalues = de_analysis[, pval_col_to_keep]
```

@@

```{r filter_genes}
if(filter_genes){
    # First, filter out any genes that doesn't have a log fold change of at
    # least 2 at at least one of the time points.

    genes_to_keep = row.names(lfc_max[rowSums(lfc_max > 2) > 0, ])
    pvalues = pvalues[genes_to_keep, ]

    if(length(genes_to_keep) > n_genes_to_keep){

	# Then rank by fisher's p-value and take max the number of genes of
	# interest
	# Filter out q-values for the pvalues table
	fishers_pval = moanin:::fisher_method(pvalues)
	fishers_qval = stats::p.adjust(fishers_pval)
        if(keep_all_significant){
	   genes_to_keep = names(fishers_qval[fishers_qval < 0.05])
	}else{
	    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
	}
    }

@@

# Rescale the centroids
centroids = moanin:::rescale_values(kmeans_clusters$centroids, meta)

# XXX is there a bug here?
# all_scores = data.frame(row.names=row.names(data))
# for(k in 1:n_clusters){
#     scores = moanin:::score_genes_centroid(data, centroids[k,])
#     all_scores[paste0("cluster_", k)] = scores
# }

# Only assign labels to X% of the genes
# max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
# genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
# labels = moanin:::row_argmin(all_scores)
# labels[genes_to_not_consider] = NA
```

@@

@@

## How to choose the number of clusters.

A common question that arises when performing clustering is how to choose the
number of clusters.
